Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-09-24 manuscript/structure/draft.qmd
Updated examples and added research context

# Abstract
The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.

# Introduction
Stochastic volatility is a type of volatility model which explicitly treats the variance as a latent random variable. These are typically expressed as non linear Gaussian state space models and are difficult to estimate using classical methods. The likelihood is unavailable in closed form and there are at least as many variables as data points. Bayesian approaches to estimating these models rely on computational Markov Chain Monte Carlo (MCMC) techniques to sample from the target joint posterior of these high dimensional latent parameter spaces.

Kim, Shephard and Chib (1998) published a method for sampling this model using a combination of Kalman Filters, Gibbs, and Metropolis Hastings over 20 years ago.  Since then, there have been many advances in Bayesian, and more broadly, statistical computation to estimate complex, high dimensional state space models. In particular, the recent availability of Hamiltonian Monte Carlo, a variant of the Metropolis Hastings algorithm, is a advancement of MCMC for efficient sampling of complex probabilistic models. The adoption of such new techniques have been made widely available through the delveopment of various open source libraries such as the Stan programming language and the PyMC library. Other algorithms have also improved the speed of estimating high dimensional models using approximations of the posterior such as Integrated nested Laplace approximation and Approximate Bayesian Computation.

These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in our statistical workfow are required to assess new sampling algorithms as well as compare computational strategies used to estimate increasingly complex models.

This research implements a simulation study to assess the algorithms used to estimate stochastic volatility models. Simulation Based Calibration outlined in Talts et al. is the primary study design to determine whether the MCMC strategy are returning efficient and well calibrated posterior estimates.

Two sampling approaches are considered in this study - Kalman Filter and Metropolis Hastings estimation by Kim Sherphard and Chib, and the Hamiltoninan Monte Carlo algorithim.

Key metrics used to evaluate the model are the effective sample size and visual tests of uniform rank statistics.

@@

# Stochastic Volatility
The model of interest in this research is the discrete time, univariate stochastic volatility model outlined by \citet{kim1998stochastic} who use Bayesian estimation methods. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

## Diagnostic limitations on real data
The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. A common set of metrics are used to check the convergence of the MCMC. This includes checking the effective sample size and the $\hat{R}$ calculations for estimated chains and parameters. These diagnostics check whether an MCMC has failed to converge onto the target posterior distribution (since we can never know with certainty that a sampler has converged).

Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when trying to diagnose the _cause_ of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either inocrrect model specification or issues in the sampling strategy (algorithm) or both.

Furthermore, different sampling strategies and MCMC tools could provide different estimates for the same model. Assuming no issues in computation, these diagnostic measures do not provide any evidence to which estimate is closer to the truth.

To illustrate this problem, suppose we estimate this model with two different sampling strategies on real data. The data are the demaned log returns of the 2019-2023 S&P 500 Index. Below are the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ and corresponding summary statistics.

@@

These particular estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$ and $\sigma^2$ possess slightly heavier right tails for SM1 compared to SM2. While in this particular example there is no major difference in the estimates, there is no certainty in whether SM1 or SM2 is consistenty closer to the truth. There are other ways of model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter _estimation_ it is unclear which one should be selected.

@@

## Diagnostic limitations on a single simulation
Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and sampling strategy on simulated data. One approach is to simulate data from the proposed model using known parameters. Then, fit the model on the simulated data and see if the model can recover the true parameters. This gives us the benefit of knowing or defining the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

Furthermore, as discussed in \cite{Bayesian workflow}, fitting models on simulated data is the only way we can check inference on latent variables. This is critical in Stochastic Volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generation which governs the latent process and reveals what the model can infer about the latent variables.

While simulations enable us to check whether our models and algorithms can capture the true generating process, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. \cite{Talts et al} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude "that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed".

@@

The posteriors for x and y sufficiently cover the true parameter, however, the true parameter for z is deep in the tails. Such an analysis may incorrectly conclude that the model fails to adequatly estimate the z parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the nuances of this specific simulated dataset.

@@

## Research Goal
The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to sample Stochastic Volatility models. Indeed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple "true parameters" are required. This methodology is discussed in the next session.

Additionally, two sampling strategies will be assessed using the proposed simulation strategy as well as the performance of the samplers under different parameterisations of the stochastic volatility model. Not only will the calibration of any one algorithm be assessed, but also compared across sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

## Metrics
The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

ESS measures the efficiency of the MCMC sampler. It calculates the number of effectively independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then we may conclude that this method is a more efficient sampler (conditional on the model).

Rank statistics as described in the SBC section are used to evaluate the calibration of a posterior. Histograms will be used to evaluate the distribution of rank statistics. If a posterior is well calibrated then it is expected that the histogram is uniform.

@@

Chi squared statistics can be used to test for uniformity and thus calibration of all the parameters in the model.

@@

## Sampling method 1
Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

@@

## Model Parameterisation
The parameterisation of a model can affect the performance of a MCMC algorithm when sampling from models with complex posterior geometries. An example of this is Neal's funnel \citep{neal2003slice} where the Hamiltonian Monte Carlo sampler encounters performance issues in hierarchical models and produces biased samples.

The stochastic volatility model as described in section 2.1 is follows a "centered" parameterisation. This describes the central location of the latent state parameters which are centered around the mean and lag of the log volatility $\mu +\phi(h_t - \mu)$. A model can be reparameterised such that the states are sampled on a distribution centered on 0 (i.e "non centered") which are later transformed to have the correct mean.  It is useful to compare model parameterisations since sampling strategies may be sensitive to the posterior geometry.

### Non centered Hamiltonian Monte Carlo
First sample from a standard normal distribuition and multiply by the variance of the log volatility.

@@

### Non centered Gaussian Mixture
The non centered model for the Gaussian approximation is expressed slightly differently due to the use of the Kalman Filter. While HMC can sample from the joint posterior directly with state vectors centered at zero, applying the Kalman Filter requires the state equation to be a function of the state variable.

@@

### Prior Predictive check
It is worth noting that the centered and non-centered parameteristions are the same stochastic volatility model. Different parameteristions express the same mathematical models in ways that may make it easier or harder for a given algorithm to sample.

@@

# Conclusion

>>>>

# Abstract
This research compares the computational methods used to estimate Bayesian stochastic volatility models. The estimation strategies outlined in the stochastic volatility literature and Hamiltoninan Monte Carlo are assesssed in their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these MCMC algorithm are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which method is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy. Results reveal that Hamiltonian Monte Carlo provides more efficient and calibrated posterior estimates conditional on the non centered paramterisation of the stochastic volatility model.

<!-- 
The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.
 -->
# Introduction
The stochastic volatility model is used in financial econometrics to model the behaviour of financial instruments. It explicitly treats the variance as a latent random variable. These are typically expressed as non linear Gaussian state space models and are difficult to estimate using classical methods. The likelihood is unavailable in closed form and there are at least as many variables as data points. Bayesian approaches to estimating these models rely on  Markov Chain Monte Carlo (MCMC) techniques to sample from the target joint posterior of these high dimensional latent parameter spaces.

Kim, Shephard and Chib (1998) propose a method for sampling this model using a combination of Kalman Filters, Gibbs sampler, and Metropolis Hastings over 20 years ago.  Since then, there have been many advances in statistical computing to estimate complex, high dimensional Bayesian models. In particular, the recent availability of Hamiltonian Monte Carlo, a variant of the Metropolis Hastings algorithm, is an advancement of MCMC for efficient sampling of complex probabilistic models. The adoption of such new techniques have been made widely available through the delveopment of various open source libraries such as the Stan programming language and the PyMC library. Other algorithms have also improved the speed of estimating high dimensional models using approximations of the posterior such as integrated nested Laplace approximation (INLA).

These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in our statistical workfow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

This research conducts a simulation study to assess the algorithms used to estimate stochastic volatility models. The first algorithm replicates Kim Sherphard and Chib's (KSC) Gaussian mixture approximation which is applied on a transformed stochastic volatility model with a log chi squared error distribution. Their sampling method uses the aforementioned Kalman Filter estimating the latent states, conjugate posterior distributions and the Metropolis Hastings algorithm. The second algorithm is the Hamiltonian Monte Carlo (HMC) algorithm as implemented in the Stan programming language. The objective is to analyse the efficiency of the sampling approach and their ability to return calibrated posterior estimates.

Comparisons of these algorithms are conducted by Simulation Based Calibration (Talts et al.). Parameters are drawn from the prior distribution and used to create datasets from the generative (stochastic volatility) model which are sampled using a MCMC method. Repeating this process multiple times gives insight to how well the sampling approach can estimate the true parameters governing the data generating process on average. The key diagnostic metrics are the effective sample size to measure the efficiency of the algorithm and visual tests of uniform rank statistics to assess calibration of the posterior estimates.

Additionally, two different parametiersations of the stochastic volatility model are considered for each sampling method. These are described as either 'centred' or 'non-centred' in location. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC efficiency as described in \citet{strickland2008parameterisation}. The key contribution of this research is to use simulations to assess whether a MCMC algorithm can return calibrated posterior estimates as well as the efficiency under the different parametiersations.

This paper is structured as follows. Section 2 provides the context around this research, namely the stochastic volatility model examined and the limitations of single simulation studies and MCMC convergence diagnostics. Section 3 describes the simulation design and diagnostic metrics. Section 4 details the sampling approaches. Section 5 discusses the results and limitations. Finally, section 6 concludes and provides points for further reserach.

 <!-- is the primary study design to determine whether the MCMC strategy are returning efficient and well calibrated posterior estimates.

Two sampling approaches are considered in this study - Kalman Filter and Metropolis Hastings estimation by Kim Sherphard and Chib, and the Hamiltoninan Monte Carlo algorithim.  -->

<!-- HMC is able to able to directly sample from the specified stochastic volatility model without any approximations of the error term.  Its key innovation is uses the gradients of the target joint posterior distribution to generate efficient proposals for the sampler to explore. -->

@@

# Stochastic Volatility
The model of interest is the discrete time, univariate stochastic volatility model estimated by \citet{kim1998stochastic} using Bayesian methods. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

## Diagnostic limitations on real data
The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. Convergence metrics are often used to check the performance of the MCMC chains, such as the effective sample size and the $\hat{R}$. These diagnostics provide evidence on whether an MCMC has failed to converge onto the target posterior distribution.

Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either inocrrect model specification or issues in the sampling algorithm or both.

Furthermore, different sampling strategies and MCMC tools could provide different posterior estimates for the same model. Assuming no issues in computation, these diagnostic measures do not provide any evidence to which estimate is closer to the truth.

To illustrate this problem, suppose we estimate this model with two different MCMC algorithms on real data. The data are the de-meaned log returns of the 2023 January-September S&P 500 Index. Below are the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ and corresponding summary statistics.

@@

These estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$  has a fatter left tail for and $\sigma^2$ has heavier right tails for sampling method 1 (SM1). Furthermore, SM2 has higher peaks at their mode relative to SM1 with a tighter spread. Based on this information alone, there is no certainty in whether SM1 or SM2 is has a more correct estimate conditional on data and model. There are other methods for model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

@@

## Diagnostic limitations on a single simulation
Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from the proposed model using known parameters. Then, fit the model on the simulated data and see if the model can recover the true parameters. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

Furthermore, as discussed in \citet{gelman2020bayesian}, fitting models on simulated data is the only way we can check inference on latent variables. This is critical in Stochastic Volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

Simulations enable us to check whether our models and algorithms can estimate the true data generating process, however, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. \cite{Talts et al} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude "that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed".

@@

The 95\% credible intervals for $\sigma$ and $\phi$ cover the true parameter. The true parameter for $\mu$ however, is deep in the tails and outside the interval. Such an analysis may incorrectly conclude that the model fails to adequatly estimate the $\mu$ parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the features of this specific simulated dataset.

@@

## Research Goal
The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to sample Stochastic Volatility models. Indeed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple "true parameters" are required. This methodology is discussed in the next section.

Additionally, two sampling strategies will be assessed using the proposed simulation design as well as the performance of the samplers under different parameterisations of the stochastic volatility model. Not only will the calibration of any one algorithm be assessed, but also compared across sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

## Evaluation Metrics
The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

### Effective sample size
ESS measures the efficiency of the MCMC sampler. It calculates the number of effectively independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then there is evidence that this method is a more efficient sampler.

### Rank statistics and chi squared statistics
Rank statistics as described in the SBC section are used to evaluate the calibration of a posterior. Histograms will be used to evaluate the distribution of rank statistics. If a posterior is well calibrated then it is expected that the histogram is uniform.

@@

A perfectly discrete uniform distribution will return a $\chi^2$ statistic of zero. That is, the number of rank statistics in each bin is equal to the expected number conditional on the number of bins and observations. Instead of performing multiple hypothesis tests for each parameter, the distribution of $\chi^2$ statistics is compared across all parameters and simulations. This will give a high level summary of calibration and overall performance of the algorithms.

<!-- % Chi squared statistics can be used to test for uniformity and thus calibration of all the parameters in the model. -->

@@

## Sampling method 1
KSC sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

@@

## Model reparameterisation
The parameterisation of a model can affect the performance of a MCMC algorithm when sampling from models with complex posterior geometries. An example of this is Neal's funnel \citep{neal2003slice} where the Hamiltonian Monte Carlo sampler encounters performance issues in hierarchical models and produces biased samples. Efficiency improvements of MCMC algorithms for state space models under different parameterisations are also explored in Forbes et al.

The stochastic volatility model as described in section 2.1 is follows a "centered" parameterisation. This describes the central location of the latent state parameters which are centered around the mean and lag of the log volatility $\mu +\phi(h_t - \mu)$. A model can be reparameterised such that the states are sampled on a distribution centered on 0 (i.e "non centered") which are later transformed to have the correct mean. The following re-parameterisations are explored as part of the simulation study.

### Non centered Hamiltonian Monte Carlo
First sample from a standard normal distribuition and multiply by the variance of the log volatility. The state variable is sampled with mean centered on zero and variance equal to $\sigma_{\eta}^2$

@@

### Non centered Gaussian Mixture
The non centered model for the Gaussian approximation is expressed slightly differently due to the use of the Kalman Filter. Non centered HMC samples from the joint posterior directly with state vectors centered at zero and transformed to return the correct log volatility estimates. Applying the Kalman Filter requires the state equation to be a function of the state variable with the average log volatility parameter $\mu$ entering the measurement equation.

@@

### Prior Predictive check
It is worth noting that the centered and non-centered parameteristions are the same stochastic volatility model. Different parameteristions express the same mathematical models in ways that make it easier or harder for a given algorithm to sample.

@@

# Conclusion
<!--

% - SV models and more broadly state space models are complex, high dimensional models which require computational tools to estimate them.
% - Modelling variance useful and important?

% - Advanced in Bayesian computation
%     + Conceptually and practically diff techniques to estimate complex models didn't exist not that long ago
%     + Performance metrics useful for checking fit on single dataset. However not much emphasis on checking performance between MCMC

% - A principled Bayesian workflow which usually requires checking of multiple componenets: prior, posterior, model choice. 
%     + Lots of different components which could contribute to an incorrect fit or error.

% - This research focuses on evaluating the computational tools used to esitmate a specific type of model.

% % One para
% - This research </contributes> examine the algos used to estimate SV model 
%     + Particular focus is on the computational tools used to estimate these models and whether they are returning what we expect.
%     + conditional on the same prior

%     - Additionally, check the sensitivity of the simulation and algortihms to the parameterisation of the model. Taking idea and advice from Catherine's paper about improvements in sampling performance.

% %  2nd 
% - Investigating two algorithms, HMC and GMA from KSC.

% % 3
% - Two samplings methods are used and are compared through simulation
%     + Specifically, the simulation designed which is taken from talts et al.

% % 4
% - Key metrics of interests are ESS for efficiency and tests of uniformity for calibration.

% - Goal is to show which algo and parameterisation is "better". Moreover, want to demonstarte that this part of a Bayesian workflow which can be applied more generally is important for checking that the model is returning the correct posteriros - and if not, where the problem may be occurring and why. Give a set of tools to evaluate algos and parametiersations (before real data is even considered). **Why is this important.**

% - Sections outlined below: -->
